CASE DETECTION OF TUBERCULOSIS PATIENTS USING MACHINE-LEARNING

### 2024-05-05 code/analysis-code/ML-analysis.qmd
part61

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","data2.rds")
#load data
# Loading the needed dataset
madaproject <- readRDS(data_location)
head(madaproject)
colnames(madaproject)
```

@@

FITTING OF ALL MODELS 
Fitting the null model
```{r}
#Fit a null model to the outcome tb
logistic_model <- logistic_reg() %>% set_engine("glm")
null_model <- logistic_model %>% fit(tb ~ 1, data = train)

@@

# Workflow (Recipe + Model)
log_lasso_wf <- workflow() %>% 
    add_recipe(logistic_rec) %>%
    add_model(logistic_lasso_spec_tune)

@@

# Tune Model (trying a variety of values of mtry and trees)
rf_grid <- grid_regular(
  mtry(range = c(1,7)), # adjusting based on number of predictors
  trees(range = c(1, 21)), # adjust the range as needed
  levels = 5 # adjust the number of levels as needed
)

tune_outputrf <- tune_grid( 
  rf_wf, # workflow
  resamples = data_cv10, # cv folds
  metrics = metric_set(roc_auc),
  control = control_resamples(save_pred = TRUE),
  grid = rf_grid, # parameter grid defined above
)
print(tune_outputrf)
# Visualize Model Evaluation Metrics from Tuning
#autoplot(tune_outputrf)

@@

# Create a flextable from the data frame
#rfconf_table <- flextable(conf_dfrf)

>>>>

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","madaproject.rds")
#load data
# Loading the needed dataset
madaproject <- readRDS(data_location)
head(madaproject)
colnames(madaproject)
```

@@

## FITTING OF ALL MODELS

Fitting the null model
```{r}
#Fit a null model to the outcome tb
logistic_model <- logistic_reg() %>% set_engine("glm")
null_model <- logistic_model %>% fit(tb ~ 1, data = train)

@@

# Workflow (Recipe + Model)
log_lasso_wf <- workflow() %>% 
    add_model(logistic_lasso_spec_tune) %>%
    dd_recipe(logistic_rec)

@@

# Setup grid tune for  Model
rf_grid <- grid_regular(
  mtry(range = c(1,7)), # adjusting based on number of predictors
  trees(range = c(1, 21)), # adjust the range as needed
  levels = 5 # adjust the number of levels as needed
)

tune_outputrf <- tune_grid( 
  rf_wf, # workflow
  resamples = data_cv10, # cv folds
  metrics = metric_set(roc_auc),
  control = control_resamples(save_pred = TRUE),
  grid = rf_grid, # parameter grid defined above
)
print(tune_outputrf)

@@

### 2024-05-05 products/manuscript/Manuscript.qmd
HardinParker Review
part61

## Introduction

Although early detection and treatment of Tuberculosis cases are the hallmark of successful TB control, diagnostic delay is still long and common in Uganda. Therefore, the study aimed atdeveloping a tool that uses machine-learning techniques to detect and diagnose TB more accurately in a shorter time.

## Methods

This was a retrospective study that used secondary data collected between 2011 to 2018 collected from health under Infectious diseases research collaboration an (IDRC) project called MIND (Mulago Inpatient Non-invasive Diagnosis for pneumonia). The study was conducted on data from 2296 tested patients of whom 1345 had no TB and 951 had TB. Different machine learning metrics(Accuracy, RMSE and AUC) were used evaluate the performance of the models.

@@

## Consclusion

The results highlight the potential of advanced machine learning algorithms, particularly Random Forest, in enhancing TB diagnosis and early detection strategies, offering valuable insights for clinical decision-making and the development of more effective screening protocols.
{{< pagebreak >}}

# Introduction

Tuberculosis (TB) remains a significant global health challenge due to diagnostic limitations affecting both pediatric and adult populations, resulting in delayed diagnosis or misdiagnosis[@sagili_strategies_2022]. This delay impacts individual prognosis and community transmission[@]. TB, caused by the bacterium Mycobacterium tuberculosis, primarily affects the lungs but can also involve other organs in its active form or remain latent without symptoms[@baykan_extrapulmonary_2022],. In 2020, TB ranked among the top 10 causes of death globally, with over 10 million new cases and 1.3 million deaths reported.
Africa bears a considerable burden, with an estimated quarter of new cases globally, resulting in approximately 417,000 deaths annually[@chakaya_who_2022]. TB is a leading cause of death among HIV-infected patients in sub-Saharan Africa, compounded by the low sensitivity of the commonly used sputum smear microscopy, particularly in detecting TB among people living with HIV[@palattiyil_access_2022].

@@

## Questions/Hypotheses to be addressed

1.  How can TB prediction be modeled using early clinical patient data?

2.  What is the performance of the TB prediction models?

# Methods

## Study Design

The study was a retrospective study. This study was based on the secondary data, which was classified as clinical examination, patients' history like alcohol use, smoking and diagnostics.

## Study Population and Setting

The study population was the patients' medical records between 2011 to 2018 who visited the hospital under an International Development Research Centre (IDRC) project. The project is currently running in China Friendship Hospital, Naguru.

## Inclusion and Exclusion Criteria

The inclusion criteria was all patients who had cough for more than two weeks and the exclusion criteria was all adults living outside Kampala.

## Data aquisition

The data was acquired from the TB clinic at Hospital patient records form patient forms and medical examination reports from the data records office. The TB dataset consisted of 2296 instances with 15 attributes.

## Data import and cleaning

Our analysis began with thorough data preparation. Following a systematic approach, we ensured data integrity and reliability by sourcing and processing the dataset from an Excel file into R. To understand the data's structure, we reviewed the codebook and conducted exploratory analyses, including summaries and visualizations. We then cleaned the data to enhance its quality for analysis. Character variables were converted to factors, and missing values were identified using visualizations and summaries and omitted. Finally, the data was strategically split into training and testing sets in an 70/30 ratio to prepare for model development and evaluation. These steps guaranteed a dataset ready for further analysis.

# Results

@@

Data analysis was conducted using R Studio version RStudio 2023.09.1. Descriptive statistics (frequencies, mean, standard deviation (SD), median, interquartile range (IQR), and proportions) were used to present the baseline characteristics of participants. we used histograms, bar graphs, and box plots, to show how data was distributed.We did cross tabulation between the covariates and the outcome.A dataset of 2294 patients was analyzed. Amongst these patients, the gender distribution was 47.0% female and 53.0% male with mean age of 31 years.

```{r echo=FALSE, message = FALSE }

ch<- readRDS(here("results", "tables", "characteristics.rds"))

ch %>%
  kable()
```

## Determining requirements for Tuberculosis prediction that is appropriate data and modeling techniques

In our study, we utilized 2296 instances and examined 14 variables associated with TB. Subsequently, nine variables were selected for prediction after conducting chi-square tests. Non-statistically significant variables associated with TB were eliminated. Following this, chi-square tests were applied to the nine chosen variables to assess their association with a threshold of 0.05. However, these tests did not reveal a significant association among the variables themselves. Therefore, they were included in the prediction model. The variables retained for prediction were: asthma, alcohol use, fever, chest radiography, home fuel use, weight loss, presence of blood in sputum, oxygen saturation, and HIV status.

```{r echo=FALSE, message = FALSE}

mpf <- readRDS(here("results", "tables", "model_perfomance_train.rds"))

mpf %>%
  kable()
```

@@

```{r echo=FALSE, message = FALSE }
rf <- readRDS(here("results", "tables", "rfconf_table.rds"))
rf %>%
  kable()
```

@@

```{r echo=FALSE, message = FALSE }
knitr::include_graphics(here("results","figures", "rf_ROCCURVE.png"))
```

```{r echo=FALSE, message = FALSE}

svm<- readRDS(here("results", "tables", "svmconf_table.rds"))
svm %>%
  kable()
```

@@

```{r echo=FALSE, message = FALSE}

knitr::include_graphics(here("results","figures", "svm_ROCCURVE.png"))
```

```{r echo=FALSE, message = FALSE}

ls<- readRDS(here("results", "tables", "lassoconf_table.rds"))
ls %>%
  kable()
```

@@

```{r echo=FALSE, message = FALSE}

knitr::include_graphics(here("results","figures", "Lasso_ROCCURVE.png"))
```

```{r echo=FALSE, message = FALSE}

tmpf <- readRDS(here("results", "tables", "model_perfomance_test.rds"))

tmpf %>%
  kable()
```

@@

The evaluation of machine learning models' performance on test data is pivotal for gauging their efficacy and generalization capabilities. The predictive accuracy of the models were measured using the test data. The Lasso, Random Forest, and Support Vector models were compared against a Null Model, with their respective performance metrics outlined. The results revealed a notable performance gap between the advanced models and the Null Model. Specifically, the Random Forest model exhibited superior performance, achieving an accuracy of 81.57% and an AUC of 0.887. In contrast, the Lasso and Support Vector models attained accuracies of 80.26% and 79.68%, with corresponding AUC values of 0.889 and 0.882, respectively.

>>>>

## Introduction
Although early detection and treatment of Tuberculosis cases are the hallmark of successful TB control, diagnostic delay is still long and common in Uganda. Therefore, the study aimed at developing a tool that uses machine-learning techniques to detect and diagnose TB more accurately in a shorter time.

## Methods
This was a retrospective study that used secondary data collected between 2011 to 2018 collected from health under Infectious diseases research collaboration an (IDRC) project called MIND (Mulago Inpatient Non-invasive Diagnosis for pneumonia). The study was conducted on data from 2296 tested patients of whom 1345 had no TB and 951 had TB. Different machine learning metrics(Accuracy, RMSE and AUC) were used evaluate the performance of the models.

@@

## Consclusion
The results highlight the potential of advanced machine learning algorithms, particularly Random Forest, in enhancing TB diagnosis and early detection strategies, offering valuable insights for clinical decision-making and the development of more effective screening protocols. {{< pagebreak >}}

# Introduction
Tuberculosis (TB) remains a significant global health challenge due to diagnostic limitations affecting both pediatric and adult populations, resulting in delayed diagnosis or misdiagnosis[@sagili_strategies_2022]. This delay impacts individual prognosis and community transmission. TB, caused by the bacterium Mycobacterium tuberculosis, primarily affects the lungs but can also involve other organs in its active form or remain latent without symptoms[@baykan_extrapulmonary_2022],. In 2020, TB ranked among the top 10 causes of death globally, with over 10 million new cases and 1.3 million deaths reported. Africa bears a considerable burden, with an estimated quarter of new cases globally, resulting in approximately 417,000 deaths annually[@chakaya_who_2022]. TB is a leading cause of death among HIV-infected patients in sub-Saharan Africa, compounded by the low sensitivity of the commonly used sputum smear microscopy, particularly in detecting TB among people living with HIV[@palattiyil_access_2022].

@@

## Questions/Hypotheses to be addressed
The prediction of TB can be modeled using early clinical patient data from Kampala residents who visited the IDRC between 2011 and 2018 and presented with a persistent cough lasting two weeks or more.

{{< pagebreak >}}
# Methods
## Study Design
The study was a retrospective study. This study was based on the secondary data, which was classified as clinical examination, patients' history like alcohol use, smoking and diagnostics.

## Study Population and Setting
The study population was the patients' medical records between 2011 to 2018 who visited the hospital under an International Development Research Centre (IDRC) project. The project is currently running in China Friendship Hospital, Naguru.

## Inclusion and Exclusion Criteria
The inclusion criteria was all patients who had cough for more than two weeks and the exclusion criteria was all adults living outside Kampala.

## Data aquisition
The data was acquired from the TB clinic at Hospital patient records form patient forms and medical examination reports from the data records office. The TB dataset consisted of 2296 instances with 15 attributes.

## Data import and cleaning
Our analysis began with thorough data preparation. Following a systematic approach, we ensured data integrity and reliability by sourcing and processing the dataset from an Excel file into R. To understand the data's structure, we reviewed the codebook and conducted exploratory analyses, including summaries and visualizations. We then cleaned the data to enhance its quality for analysis. Character variables were converted to factors, and missing values were identified using visualizations and summaries and omitted. Finally, the data was strategically split into training and testing sets in an 70/30 ratio to prepare for model development and evaluation. These steps guaranteed a dataset ready for further analysis.

## Model Development
Three models were considered for the prediction analysis, and each were fit for both outcomes of interest (tuberculosis). The data were split such that 70% were used for training the models and the remaining 30% were used for testing the final chosen model. Cross-validation was used for all models, using a ten-fold resampling structure. The tidymodel framework was utilized in the analysis.

## Model Definition

There are several machine learning models suitable for addressing the research question, but this analysis primarily focused on three different types of models: Random Forest, LASSO, and Support Vector Machine (SVM).

Random Forest (RF): RF is a versatile and robust ensemble learning method that aims to reduce variance by building multiple decision trees on random subsets of the data and then averaging their predictions. It can handle large datasets with high-dimensional feature spaces and is less prone to overfitting compared to individual decision trees[@simon_interpreting_2023]. However, RF models are more complex and may be challenging to interpret compared to simpler models like decision trees.

LASSO: LASSO is a regularization-based linear regression method that balances model complexity and goodness of fit by penalizing the absolute values of the regression coefficients. This encourages sparsity in the model, meaning it tends to select a subset of the most important features while shrinking the coefficients of less important ones towards zero[@zhao_model_2006]. However, LASSO may struggle with highly correlated predictors and can only select one variable from a group of highly correlated variables, potentially ignoring important information.

Support Vector Machine (SVM): SVM is a powerful supervised learning algorithm commonly used for classification tasks. It works by finding the optimal hyperplane that separates data points belonging to different classes with the maximum margin of separation. SVM is effective in high-dimensional spaces and is versatile in handling linear and non-linear relationships through different kernel functions.[@javatpoint_support_2021] However, SVM can be computationally intensive, especially with large datasets, and may require careful tuning of hyperparameters for optimal performance.

These three models were selected to provide a diverse set of approaches for predicting TB classification based on early clinical patient data. Random Forest offers a robust ensemble learning approach, LASSO provides a sparse and interpretable linear model, and SVM offers flexibility in handling complex relationships between predictors. By comparing and contrasting the performance of these models, we aim to identify the most effective approach for TB prediction in our study population. {{< pagebreak >}}

## Model Performance Evaluation

A confusion matrix or contingency table was used to determine the classification performance of the various classification models. Given m classes, the confusion matrix is a table of at least size m by m [@riehl_hierarchical_2023]. In a confusion matrix, if the instance is positive and it is classified as positive, it is counted as a true positive (TP); if it is classified as negative, it is counted as a false negative (FN). If the instance is negative and it is classified as negative, it is counted as a true negative (TN); if it is classified as positive, it is counted as a false positive (FP). These terms are useful when analyzing a classifier’s ability to determine accuracy based on the correct and incorrect classes produced by the confusion matrix. Accuracy was used as it signifies the proportion of the total number of TB patient predictions that are correct[@carvalho_machine_2019]. True positive means the number of actual TB patients that are accurately classified, and true negatives refer to the number of patients that are correctly classified as not having TB. The ROC (Receiver Operating Characteristic) curve was also used, which is a graphical representation to measure the performance of the classifier system. It shows the trade-off between sensitivity and specificity for every possible cut-off for a test[@robinson_validating_2020]. The area under the curve ranges from 0 to 1, where 1 implies an excellent test and 0 implies a useless test.
{{< pagebreak >}}
# Results

@@

Data analysis was conducted using R Studio version RStudio RStudio 2024.04.0. Descriptive statistics (frequencies, mean, standard deviation (SD), median, interquartile range (IQR), and proportions) were used to present the baseline characteristics of participants. we used histograms, bar graphs, and box plots, to show how data was distributed.We did cross tabulation between the covariates and the outcome.A dataset of 2294 patients was analyzed. Amongst these patients, the gender distribution was 47.0% female and 53.0% male with mean age of 31 years.

## Model results and accuracy assessment

```{r echo=FALSE, message = FALSE}
#| label: tbl-schematic1
#| tbl-cap: "Model Perfomance for the Training Dataset"
mpf <- readRDS(here("results", "tables", "model_perfomance_train.rds"))

mpf %>%
  knitr::kable()
```

@@

```{r echo=FALSE, message = FALSE }
#| label: tbl-schematic2
#| tbl-cap: "Random Forest"
rf <- readRDS(here("results", "tables", "rfconf_table.rds"))
rf %>%
  knitr::kable()
```

@@

```{r echo=FALSE, message = FALSE }
#| label: fig-schematic1
#| fig-cap: "Random Forest ROC Curve"
#| echo: false
knitr::include_graphics(here("results","figures", "rf_ROCCURVE.png"))
```

```{r echo=FALSE, message = FALSE}
#| label: tbl-schematic5
#| tbl-cap: "Support Vector Machine"
svm<- readRDS(here("results", "tables", "svmconf_table.rds"))
svm %>%
  knitr::kable()
```

@@

```{r echo=FALSE, message = FALSE}
#| label: fig-schematic2
#| fig-cap: "Support Vector Machine ROC Curve"
#| echo: false
knitr::include_graphics(here("results","figures", "svm_ROCCURVE.png"))
```

```{r echo=FALSE, message = FALSE}
#| label: tbl-schematic3
#| tbl-cap: "lassoconf_table"
ls<- readRDS(here("results", "tables", "lassoconf_table.rds"))
ls %>%
  knitr::kable()
```

@@

```{r echo=FALSE, message = FALSE}
#| label: fig-schematic3
#| fig-cap: "LASSO ROC Curve"
#| echo: false 
knitr::include_graphics(here("results","figures", "Lasso_ROCCURVE.png"))
```

```{r echo=FALSE, message = FALSE}
#| label: tbl-schematic4
#| tbl-cap: "Model Perfomance Testing Results"
tmpf <- readRDS(here("results", "tables", "model_perfomance_test.rds"))

tmpf %>%
 knitr:: kable()
```

@@

The evaluation of machine learning models' performance on test data is pivotal for gauging their efficacy and generalization capabilities. The predictive accuracy of the models were measured using the test data. The Lasso, Random Forest, and Support Vector models were compared against a Null Model, with their respective performance metrics outlined. The results revealed a notable performance gap between the advanced models and the Null Model. Specifically, the Random Forest model exhibited superior performance, achieving an accuracy of 81.57% and an AUC of 0.887. In contrast, the Lasso and Support Vector models attained accuracies of 80.26% and 79.68%, with corresponding AUC values of 0.889 and 0.882, respectively.
{{< pagebreak >}}
